A simplified dynamic bioenergetic model for coral-\emph{Symbiodinium} symbioses and coral bleaching as an alternate stable state

### 2017-01-11 JTB_ms/JTB_ms.tex
finish draft

Simulations using the discretized scheme were performed using R code
that is available in the data repository accompanying this article:
github.com/jrcunning/Rcoral. By experimentation, we found that a time
step of 0.1 days gave adequate precision for most simulations (including
used to generate Figs. XXX in this paper). For steady state estimations,
simulations were run until the change in specific growth rate of host
and symbiont and the \(S:H\) biomass ratio was less than 1e-5 per time
step. In regions of state space where very slow transient dynamics could
be expected (i.e.~near bifurcation points), sample steady state
calculations were verified using MATHEMATICA code for numerical root
finding (function FindRoot) with the code written independently by a
coauthor without reference to the R code.

@@

\begin{figure}[htbp]
\centering
\includegraphics{../img/Fig6.png}
\caption{Coral bleaching as a switch from a nitrogen- to carbon-limited
alternative stable state. This transition is demonstrated in response to
gradually increasing external light (A), which causes an increase in
production of ROS (B) that reduces the photosynthetic rate through
photoinhibition (C). Decreasing photosynthesis moves the system from
overall nitrogen-limitation toward carbon-limitation (D); when this
threshold is crossed, the system rapidly becomes highly carbon-limited
as photosynthesis becomes CO\textsubscript{2}-limited (E) and symbiont
densities rapidly decline (F) into a bleached state. (All parameters at
default values; external DIN=1e-7 molN/L, prey density=0).}
\end{figure}

@@

Since bleaching can be understood as a transition from a
nitrogen-limited state with high S:H biomass to a carbon-limited state
with low S:H biomass, induced by an external stressor, recovery can be
understood as a switch back to the nitrogen-limited state once the
external stressor is alleviated. In natural settings, this typically
occurs through seasonal declines in temperature and light. However,
hysteresis associated with the system's alternate stable states
indicates that the symbiosis cannot recover along the same trajectory it
followed during bleaching; indeed, the stressor must be alleviated
\emph{below} the threshold that initially caused bleaching in order for
the system to recover (Fig. 2). This is because under the same external
conditions, a bleached coral with low S:H biomass (relative to a healthy
coral with high S:H biomass) is characterized by greater light
amplification and weaker CCM activity, which enhance photooxidative
stress and serve to maintain the carbon-limited state. In order for the
system to recover, the stressor must be reduced enough such that
photooxidative stress ceases and translocation of carbon from symbiont
to host is resumed. Once this occurs, the host can energize its CCMs,
which further enhances carbon fixation and accelerates the system back
toward a nitrogen-limited state, indicative of recovery. The conditions
under which recovery can occur -- which determine the magnitude of
hysteresis (Fig. SX) -- depend on the relative abundance of nitrogen and
carbon in the environment. Higher food levels, representing a
non-autotrophic carbon source for the host, make it easier for the host
to overcome carbon-limitation, thus providing a potential mechanism
underlying observations that feeding aids recovery from bleaching
{[}xxx{]}. Conversely, high external DIN impedes the re-establishment of
nitrogen-limitation, making recovery from bleaching more difficult
(i.e., narrowing the sets of conditions under which recovery is
possible).

Dynamic simulations of recovery reveal another interesting behavior of
the system: under some conditions, an `overshoot' occurs in which S:H
biomass temporarily increases beyond the ratio maintained in the
`healthy' state, before returning to stabilize at this value (Fig. 8;
Fig. S9). In fact, unusually high symbiont densities have been observed
following bleaching in both experimental (Cunning, Silverstein, and
Baker 2015) and field studies (Kemp et al. 2014), and have been
interpreted as a potential `disequilibrium in host-symbiont regulation'
(Kemp et al. 2014). The model reveals the dynamics of this `overshoot'
as follows: as symbionts repopulate the host, photosynthesis becomes
increasingly CO\textsubscript{2}-limited due to weak CCM activity of the
carbon-limited host. Thus, although symbiont growth is not yet
carbon-limited, a growing symbiont population has less and less excess
carbon (per symbiont) to share, and is thus moving toward
carbon-limitation. Meanwhile, because S:H biomass is increasing, the
host receives more and more carbon per unit host biomass, and is thus
moving away from carbon-limitation. If the host overcomes
carbon-limitation \emph{before} the symbiont reaches it, then the system
rapidly transitions to the nitrogen-limited state and the S:H ratio
stabilizes without an overshoot (Fig. 8A). However, if the symbiont
becomes carbon-limited first (Fig. 8B, 8C), then carbon translocation
per symbiont declines further and photosynthesis becomes more
CO\textsubscript{2}-limited, which maintains carbon-limitation of the
host. In this situation, continued growth of less and less productive
symbionts causes S:H biomass to reach a much higher level before the
host finally overcomes carbon-limitation. At this point, representing
the peak of the overshoot, the transition to nitrogen-limitation finally
takes place and the S:H biomass ratio declines and stabilizes.

@@

\begin{figure}[htbp]
\centering
\includegraphics{../img/Fig8.png}
\caption{Recovery from bleaching with varying N:C availability (A:
N:C=0.09; B: N:C=0.10; C: N:C=0.30). Higher N:C ratios in a
heterotrophic food source (effectively representing higher external DIN
and/or lower feeding rates) cause a larger overshoot in the S:H biomass
ratio, and prolong the duration of time until the system `recovers' by
re-establishing nitrogen-limitation. All simulations run with default
parameters (except varying nNX), with L=10 mol photons
m\textsuperscript{2} d\textsuperscript{-1}, DIN=1e-7 molN
L\textsuperscript{-1}, prey=1e-6 CmolX L\textsuperscript{-1}, and
initial S:H biomass=0.001.}
\end{figure}

@@

This simplified dynamic bioenergetic model of coral-\emph{Symbiodinium}
symbioses mechanistically reproduces patterns in steady-state coral
growth and symbiont abundance commonly observed in corals, including
higher symbiont abundance with higher nutrients and feeding, lower
symbiont abundance with increasing light, and optimal growth at
intermediate light levels. Moreover, the model reproduces complex
dynamic behaviors including seasonal changes in symbiont density at
different nutrient levels, rapid bleaching above a threshold of high
light, mitigation of bleaching by heterotrophic feeding, exacerbation of
bleaching by elevated nutrients, and an overshoot of symbiont density
during recovery from bleaching. These examples demonstrate the model's
ability to integrate multiple environmental forcing functions to
reproduce complex responses; meanwhile, the diversity of these phenomena
suggest the model has captured many of the important features of the
system in a unifying mechanistic framework. This model also provides a
new conceptual framework for considering coral bleaching as a transition
to an alternate stable state, which has important implications for
understanding the performance and maintenance of symbiotic interactions.
In this context, the `healthy' stable state represents a scenario in
which the host is `in control' of the system, with nitrogen-limitation
stabilizing the symbiont to host biomass ratio and maintaining positive
growth. Conversely, carbon-limitation represents a loss of host
`control', wherein positive feedbacks result in a loss of capacity for
photosynthetic carbon fixation and positive growth. Importantly, these
`control' mechanisms arise entirely passively through feedbacks present
in the system.

In addition to the suite of specific examples presented here, the model
can be used to explore many different dynamic environmental scenarios,
and represents a tool that biologists and ecologists can use to generate
hypotheses and make predictions in both experimental and natural
settings. Moreover, parameter values can be modified to correspond to
different genetic or functional types of coral hosts and
\emph{Symbiodinium} partners in order to evaluate variability in system
responses. Thus, the diversity of potential applications for this model
is high, and we envision this work as a foundation for continued
development, which may include more detailed treatments of specific
modules (e.g., DIC processing), and the incorporation of more external
forcing capabilities (e.g., external DIC). Importantly, this effort is
best undertaken by the scientific community at large, including those
with theoretical and empirical backgrounds, which is why the code for
this model has been developed openly using the R language. Ultimately,
the continued refinement of these tools is fundamental in elucidating
the mechanisms of symbiosis function and dysfunction, and in predicting
coral responses to environmental change.

@@

>>>>

Simulations using the discretized scheme were performed using R code
that is available in the data repository accompanying this article:
github.com/jrcunning/Rcoral. By experimentation, we found that a time
step of 0.1 days gave adequate precision for most simulations (including
used to generate Figs. 2-8 in this paper). For steady state estimations,
simulations were run until the change in specific growth rate of host
and symbiont and the S:H biomass ratio was less than 1e-5 per time step.
In regions of state space where very slow transient dynamics could be
expected (i.e.~near bifurcation points), sample steady state
calculations were verified using MATHEMATICA code for numerical root
finding (function FindRoot) with the code written independently by a
coauthor without reference to the R code.

@@

\begin{figure}[htbp]
\centering
\includegraphics{../img/Fig6.png}
\caption{Coral bleaching as a switch from a nitrogen- to carbon-limited
alternative stable state. This transition is demonstrated in response to
gradually increasing external light (A), which causes an increase in
production of ROS (B) that reduces the photosynthetic rate through
photoinhibition (C). Decreasing photosynthesis moves the system from
overall nitrogen-limitation toward carbon-limitation (D); when this
threshold is crossed, the system rapidly becomes highly carbon-limited
as photosynthesis becomes CO\textsubscript{2}-limited (E) and symbiont
densities rapidly decline (F) into a bleached state. Substrate
limitation coefficients were calculated using Eq. 22. All parameters
were set at default values with external DIN=1e-7 molN/L and prey
density=0).}
\end{figure}

@@

Since bleaching can be understood as a transition from a
nitrogen-limited state with high S:H biomass to a carbon-limited state
with low S:H biomass, induced by an external stressor, recovery can be
understood as a switch back to the nitrogen-limited state once the
external stressor is alleviated. In natural settings, this typically
occurs through seasonal declines in temperature and light. However,
hysteresis associated with the system's alternate stable states
indicates that the symbiosis cannot recover along the same trajectory it
followed during bleaching; indeed, the stressor must be alleviated
\emph{below} the threshold that initially caused bleaching in order for
the system to recover (Fig. 2). This is because under the same external
conditions, a bleached coral with low S:H biomass (relative to a healthy
coral with high S:H biomass) is characterized by greater light
amplification and weaker CCM activity, which enhance photooxidative
stress and serve to maintain the carbon-limited state. In order for the
system to recover, the stressor must be reduced enough such that
photooxidative stress ceases and translocation of carbon from symbiont
to host is resumed. Once this occurs, the host can energize its CCMs,
which further enhances carbon fixation and accelerates the system back
toward a nitrogen-limited state, indicative of recovery. The conditions
under which recovery can occur -- which determine the magnitude of
hysteresis (Fig. S9) -- depend on the relative abundance of nitrogen and
carbon in the environment. Higher food levels, representing a
non-autotrophic carbon source for the host, make it easier for the host
to overcome carbon-limitation, thus providing a potential mechanism
underlying observations that feeding aids recovery from bleaching
(Grottoli, Rodrigues, and Palardy 2006; Connolly, Lopez-Yglesias, and
Anthony 2012). Conversely, high external DIN impedes the
re-establishment of nitrogen-limitation, making recovery from bleaching
more difficult (i.e., narrowing the sets of conditions under which
recovery is possible).

Dynamic simulations of recovery reveal another interesting behavior of
the system: under some conditions, an `overshoot' occurs in which S:H
biomass temporarily increases beyond the ratio maintained in the
`healthy' state, before returning to stabilize at this value (Fig. 8).
In fact, unusually high symbiont densities have been observed following
bleaching in both experimental (Cunning, Silverstein, and Baker 2015)
and field studies (Kemp et al. 2014), and have been interpreted as a
potential `disequilibrium in host-symbiont regulation' (Kemp et al.
2014). The model reveals the dynamics of this `overshoot' as follows: as
symbionts repopulate the host, photosynthesis becomes increasingly
CO\textsubscript{2}-limited due to weak CCM activity of the
carbon-limited host. Thus, although symbiont growth is not yet
carbon-limited, a growing symbiont population has less and less excess
carbon (per symbiont) to share, and is thus moving toward
carbon-limitation. Meanwhile, because S:H biomass is increasing, the
host receives more and more carbon per unit host biomass, and is thus
moving away from carbon-limitation. If the host overcomes
carbon-limitation \emph{before} the symbiont reaches it, then the system
rapidly transitions to the nitrogen-limited state and the S:H ratio
stabilizes without an overshoot (Fig. 8A). However, if the symbiont
becomes carbon-limited first (Fig. 8B, 8C), then carbon translocation
per symbiont declines further and photosynthesis becomes more
CO\textsubscript{2}-limited, which maintains carbon-limitation of the
host. In this situation, continued growth of less and less productive
symbionts drives S:H biomass to a much higher level before the host
finally overcomes carbon-limitation. At this point, representing the
peak of the overshoot, the transition to nitrogen-limitation finally
takes place and the S:H biomass ratio declines and stabilizes as
positive growth is resumed.

@@

\begin{figure}[htbp]
\centering
\includegraphics{../img/Fig8.png}
\caption{Recovery from bleaching with varying N:C availability (A:
N:C=0.09; B: N:C=0.10; C: N:C=0.30). Higher N:C ratios in a
heterotrophic food source (effectively representing higher external DIN
and/or lower feeding rates) cause a larger overshoot in the S:H biomass
ratio, and prolong the duration of time until the system `recovers' by
re-establishing nitrogen-limitation. Substrate limitation coefficients
were calculated using Eq. 22. All simulations were run with default
parameters (except varying \(n_{NX}\)), with L=10 mol photons
m\textsuperscript{2} d\textsuperscript{-1}, DIN=1e-7 molN
L\textsuperscript{-1}, prey=1e-6 CmolX L\textsuperscript{-1}, and
initial S:H biomass=0.001.}
\end{figure}

@@

This simplified dynamic bioenergetic model of coral-\emph{Symbiodinium}
symbioses mechanistically reproduces patterns in steady-state coral
growth and symbiont abundance commonly observed in corals, including
higher symbiont abundance with higher nutrients and feeding, lower
symbiont abundance with increasing light, and optimal growth at
intermediate light levels. Moreover, the model reproduces complex
dynamic behaviors including seasonal changes in symbiont density at
different nutrient levels, rapid bleaching above a threshold of high
light, mitigation of bleaching by heterotrophic feeding, exacerbation of
bleaching by elevated nutrients, and an overshoot of symbiont density
during recovery from bleaching. These examples demonstrate the model's
ability to integrate multiple environmental forcing functions to
reproduce complex responses; meanwhile, the diversity of these phenomena
suggest the model has captured many of the important features of the
system in a unifying mechanistic framework. This model also provides a
new conceptual framework for considering coral bleaching as a transition
to an alternate stable state, which has important implications for
understanding the performance and maintenance of symbiotic interactions.
In this context, the `healthy' stable state represents a scenario in
which nitrogen-limitation stabilizes the symbiont to host biomass ratio
and maintains positive growth. Conversely, carbon-limitation represents
a dysfunctional state wherein positive feedbacks result in a loss of
photosynthetic capacity and negative growth. Importantly, these
`control' mechanisms arise entirely passively through feedbacks present
in the system.

In addition to the suite of specific examples presented here, the model
can be used to explore many different dynamic environmental scenarios,
and represents a tool that biologists and ecologists can use to generate
hypotheses and make predictions in both experimental and natural
settings. Moreover, parameter values can be modified to correspond to
different genetic or functional types of coral hosts and
\emph{Symbiodinium} partners in order to evaluate variability in system
responses. Thus, the diversity of potential applications for this model
is high, and we envision this work as a foundation for continued
development, which may include more detailed treatments of specific
modules (e.g., DIC processing), and the incorporation of more external
forcing capabilities (e.g., external DIC, temperature). Importantly,
this effort may be undertaken by the scientific community at large,
including those with theoretical and empirical backgrounds, which is why
the code for this model has been developed openly using the R language.
Ultimately, the continued refinement of these tools is fundamental in
elucidating the mechanisms of symbiosis function and dysfunction, and in
predicting coral responses to environmental change.

@@

\hypertarget{ref-Grottoli:2006p3692}{}
Grottoli, Andréa G, Lisa J Rodrigues, and James E Palardy. 2006.
``Heterotrophic plasticity and resilience in bleached corals.''
\emph{Nature} 440 (7088): 1186--9.
doi:\href{https://doi.org/10.1038/nature04565}{10.1038/nature04565}.
